---
title: "HIV Retention & Disengagement Prediction"
author: "USAID | ADAT "
date: "May 20, 2022"
output:
  html_document:
    toc: true
    toc_depth: 5
    number_sections: false
---



# Premise

Despite the progress in HIV care in suppressing viral load, disengagement from HIV care remains a significant issue that impairs the 
path of achieving the global target to end the HIV/AIDS epidemic by 2030, set forth by WHO and the Joint United Nations Programme on HIV/AIDS (UNAIDS). In light of the above, we sought to develop and validate data-driven / AI rules than can be used to foster the early identification of patients at risk of disengagement from care.

# Objective 

> The main objective of this study is to predict disengagement by one month denoted by y_2. 

# Training & Validation of Models

We leveraged Super Learner (Stacked Ensamble) algorithm to train and validated ML models. The details of implementation can be found here:

https://rpubs.com/akimaina/hiv-disengagement-prediction



# Step 1: Install / Load Relevant Packages

```{r fig.align='center',warning=FALSE, cache=F, echo=T}


# function to install missing packages
ipak <- function(pkg){
  new.pkg <- pkg[!(pkg %in% installed.packages()[, "Package"])]
  if (length(new.pkg))
    install.packages(new.pkg, dependencies = TRUE, repos='http://cran.rstudio.com/')
  sapply(pkg, require, character.only = TRUE)
}

#install.packages('package_name', dependencies=TRUE, repos='http://cran.rstudio.com/')

packages =c(  "dplyr", "tidyverse", "highcharter", "zoo", "runner", "readr", "h2o")

ipak(packages)

select = dplyr::select; summarize = dplyr::summarize; rename = dplyr::rename; mutate = dplyr::mutate;

```

‘h2o’ version 3.36.1.2 was used

# Step 2: Intialize H2o Cluster & load helper function


```{r fig.align='center',warning=FALSE, cache=F, echo=T}

# Set seed
set.seed(123)

# init h20
h2o.init( ip = "localhost", port = 7070, max_mem_size = "16G", nthreads = -1)

# Load helper functions
source("helpers.R") #

```

# Step 3. Data Pre-processing



```{r fig.align='center', cache=T, fig.width=10, warning=FALSE}
raw.df =  readr::read_csv("../../data/synthetic_data_mimics_output_of_SQL_query.csv", na = "NULL")
k_folds=6 # please not that the last fold will be used as a proxy for external validation set
clean.long.df = clean_longitudinal_data(raw.df)
clean.df= clean.long.df %>%
    filter(Enrollment_Date >= as.Date("2016-01-01") & RTC_Date <= as.Date("2023-01-01")) %>%
    assign_folds(k_folds) #%>% group_by(patientID) %>% arrange( patientID, desc(Visit_Number)) %>% filter(row_number()==4)%>%ungroup() 
```



# Step 4. Handle Missing Data 
 
```{r fig.align='center',warning=FALSE, cache=F, echo=T}
sigma=1
# Handle Missing Data
model.df = clean.df%>%
  mutate(
    #Patient_Care_Status=as.factor(Patient_Care_Status),
    y0=as.factor(`disengagement-1day`),
    y0_weight = if_else(y0=='Disengaged', (sum(clean.df[,"disengagement-1day"] == 'Active In Care', na.rm=T) / nrow(clean.df))*sigma,
                        1-sum(clean.df[,"disengagement-1day"] == 'Active In Care', na.rm=T) / nrow(clean.df)),
    y1=as.factor(`disengagement-2wks`),
    y1_weight = if_else(y1=='Disengaged', (sum(clean.df[,"disengagement-2wks"] == 'Active In Care', na.rm=T) / nrow(clean.df))*sigma,
                        1-sum(clean.df[,"disengagement-2wks"] == 'Active In Care', na.rm=T) / nrow(clean.df)),
    
    y2=as.factor(`disengagement-1month`),
    y2_weight = if_else(y2=='Disengaged', (sum(clean.df[,"disengagement-1month"] == 'Active In Care', na.rm=T) / nrow(clean.df))*sigma,
                        1-sum(clean.df[,"disengagement-1month"] == 'Active In Care', na.rm=T) / nrow(clean.df)),
    y3=as.factor(`disengagement-3month`),
    y3_weight = if_else(y3=='Disengaged', (sum(clean.df[,"disengagement-3month"] == 'Active In Care', na.rm=T) / nrow(clean.df))*sigma,
                        1-sum(clean.df[,"disengagement-3month"] == 'Active In Care', na.rm=T) / nrow(clean.df)),
    
    y4=as.factor(`disengagement-7days`),
    y4_weight = if_else(y4=='Disengaged', (sum(clean.df[,"disengagement-7days"] == 'Active In Care', na.rm=T) / nrow(clean.df))*sigma,
                        1-sum(clean.df[,"disengagement-7days"] == 'Active In Care', na.rm=T) / nrow(clean.df))
  )%>%
  mutate(
    
    # Add missing Data Indicators
    Age_NA= ifelse(is.na(Age), 1, 0),
    Duration_in_HIV_care_NA= ifelse(is.na(Duration_in_HIV_care),  1, 0),
    Weight_NA= ifelse(is.na(Weight), 1, 0),
    Height_NA= ifelse(is.na(Height), 1, 0),
    BMI_NA= ifelse(is.na(BMI), 1, 0),
    WHO_staging_NA= ifelse(is.na(WHO_staging),  1, 0),
    Regimen_Line_NA= ifelse(is.na(Regimen_Line),  1, 0),
    HIV_disclosure_NA= ifelse(is.na(HIV_disclosure),  1, 0),
    CD4_NA= ifelse(is.na(CD4), 1, 0),
    Viral_Load_log10_NA= ifelse(is.na(Viral_Load_log10), 1, 0),
    #Days_to_Start_of_ART_NA= ifelse(is.na(Days_to_Start_of_ART),  1, 0), #REMOVED
    Adherence_Counselling_Sessions_NA = ifelse(is.na(Adherence_Counselling_Sessions),  1, 0),
    Days_defaulted_in_prev_enc_NA= ifelse(is.na(Days_defaulted_in_prev_enc), 1, 0),
    num_2wks_defaults_last_3visits_NA= ifelse(is.na(num_2wks_defaults_last_3visits), 1, 0),
    ever_defaulted_by_1m_in_last_1year_NA= ifelse(is.na(ever_defaulted_by_1m_in_last_1year), 1, 0),
    ever_defaulted_by_1m_in_last_2year_NA= ifelse(is.na(ever_defaulted_by_1m_in_last_2year), 1, 0),
    
    # Set NA to a fixed value (usually zero)
    Age= ifelse(is.na(Age), 0, Age),
    Duration_in_HIV_care= ifelse(is.na(Duration_in_HIV_care), 0, Duration_in_HIV_care),
    Weight= ifelse(is.na(Weight), 0, Weight),
    Height= ifelse(is.na(Height),0, Height),
    BMI= ifelse(is.na(BMI),0, BMI),
    WHO_staging= ifelse(is.na(WHO_staging), 0, WHO_staging),
    Regimen_Line= ifelse(is.na(Regimen_Line), 0, Regimen_Line),
    HIV_disclosure= ifelse(is.na(HIV_disclosure), 0, HIV_disclosure),
    CD4= ifelse(is.na(CD4),0, CD4),
    Viral_Load_log10= ifelse(is.na(Viral_Load_log10), log10(1), Viral_Load_log10),
    #Days_to_Start_of_ART= ifelse(is.na(Days_to_Start_of_ART), 0, Days_to_Start_of_ART),  #REMOVED
    Adherence_Counselling_Sessions_NA= ifelse(is.na(Adherence_Counselling_Sessions), 0, Adherence_Counselling_Sessions),
    Days_defaulted_in_prev_enc= ifelse(is.na(Days_defaulted_in_prev_enc),0, Days_defaulted_in_prev_enc),
    num_2wks_defaults_last_3visits= ifelse(is.na(num_2wks_defaults_last_3visits),0, num_2wks_defaults_last_3visits),
    ever_defaulted_by_1m_in_last_1year= ifelse(is.na(ever_defaulted_by_1m_in_last_1year),0,ever_defaulted_by_1m_in_last_1year),
    ever_defaulted_by_1m_in_last_2year= ifelse(is.na(ever_defaulted_by_1m_in_last_2year),0, ever_defaulted_by_1m_in_last_2year)
   
    
  )

```



# Step 5. Define Predictors


```{r fig.align='center',warning=FALSE, cache=F, echo=T}

# predictors
x=c(    'Age','Age_NA', 
        'Gender' ,  
        'Marital_status', 
        'Duration_in_HIV_care', 'Duration_in_HIV_care_NA',  
        'BMI', 'BMI_NA',
        'Travel_time',  
        #'Days_to_Start_of_ART', 'Days_to_Start_of_ART_NA', 
        'WHO_staging','WHO_staging_NA',
        'Viral_Load_log10', 'Viral_Load_log10_NA', 'VL_suppression', 'Days_Since_Last_VL',
        'HIV_disclosure','HIV_disclosure_NA', 
        'Regimen_Line', 'Regimen_Line_NA',  
        'Pregnancy',
        'Clinic_Location',
        'TB_Comorbidity',
        'CD4','CD4_NA', 'Days_Since_Last_CD4',
        'Entry_Point', 
        "Encounter_Type_Class",
         'Education_Level',  
        "Occupation",
       # "Presence_of_OIs", 
        "Adherence_Counselling_Sessions",
        "Clinic_Name", 
        'ART_regimen',
        'Visit_Number', 
        'Days_defaulted_in_prev_enc', 'Days_defaulted_in_prev_enc_NA',
         'num_2wks_defaults_last_3visits', 'num_2wks_defaults_last_3visits_NA',
        'ever_defaulted_by_1m_in_last_1year','ever_defaulted_by_1m_in_last_1year_NA',
         'ever_defaulted_by_1m_in_last_2year','ever_defaulted_by_1m_in_last_2year_NA',
        
        # Baseline => Added in V3
        'Age_baseline',
        'Gender_baseline' ,  
        'Marital_status_baseline', 
        'BMI_baseline',
        'Travel_time_baseline',  
        'WHO_staging_baseline',
        'VL_suppression_baseline', 
        'Viral_Load_log10_baseline',
        'HIV_disclosure_baseline',
        'Regimen_Line_baseline', 
        'Pregnancy_baseline',
        'Clinic_Location_baseline',
        'TB_Comorbidity_baseline',
        'CD4_baseline',
        'Education_Level_baseline',  
        "Occupation_baseline",
       # "Presence_of_OIs_baseline", 
        "Adherence_Counselling_Sessions_baseline",
        "Clinic_Name_baseline", 
        'ART_regimen_baseline'
)

```


 

# Step 6. Define ML algorithm & Hyper-parameter tuning

```{r fig.align='center', fig.width=6, warning=FALSE, cache=F, results='hide', echo=T}

# Experimentation
hyper_parameters = list(
    max_models=2,#200 #10
    stopping_rounds = 3,#15 #10
    stopping_tolerance = 0.01,#0.0000001#0.0001 #0.02
    fold_column <- "fold_id",
    k_folds = k_folds,
    max_runtime_mins = 3,#22000000#80
    balance_classes = F, # false by default
    include_algos= c(
      "XGBoost",#"XRF",
      "StackedEnsemble",  "GLM",
    "GBM"
    
     ),
    stopping_metric="logloss",
    sort_metric="auc",
    assignment_type="stratified",
    verbosity= NA
    
)
  
```




# 1 day Disengagement | Weighted

## Train the model
```{r fig.align='center',warning=FALSE, cache=F}
y='y0'
weights_column = 'y0_weight'
export_path = "training scripts/V5/exports/y0_1days_IIT"
autoML.y0 <- train_ml_model(model.df, y, x, hyper_parameters, export_path)
```


## Test the model

```{r fig.align='center', cache=F, fig.width=6, warning=FALSE}

evaluate_ml_model(main.df, y, x, hyper_parameters, autoML.y0)

```










# 7 day Disengagement | Weighted

## Train the model
 
```{r fig.align='center',warning=FALSE, cache=F}
y='y4'
x = c(x)
weights_column = 'y4_weight'
export_path = "training scripts/V5/exports/y4_7days_IIT"
autoML.y4 <- train_ml_model(model.df, y, x, hyper_parameters, export_path)

```






## Test the model

```{r fig.align='center', cache=F, fig.width=6, warning=FALSE}

evaluate_ml_model(main.df, y, x, hyper_parameters, autoML.y4)

```





# 90 day Disengagement | Weighted

## Train the model

```{r fig.align='center',warning=FALSE, cache=F}
y='y3'
weights_column = 'y3_weight'
export_path = "training scripts/V5/exports/y3_90days_IIT"
autoML.y3 <- train_ml_model(model.df, y, x, hyper_parameters, export_path)

```

## Test the model

```{r fig.align='center', cache=F, fig.width=6, warning=FALSE}

evaluate_ml_model(main.df, y, x, hyper_parameters, autoML.y4)

```








